---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# FAMoS

FAMoS provides an automated and unbiased model selection algorithm that aims at determining the most appropriate subset of model parameters to describe a specific data set. It only needs three inputs:

- A vector containing all parameters
- A cost function that calculates the negative log-likelihood for a given parameter set
- The number of data points used for analysis

Due to its flexibility with respect to the cost function, the FAMoS can be used for nonlinear problems such as ordinary and partial differential equations.

## Installation

You can install FAMoS from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("GabelHub/FAMoS")
# alternative installation command
devtools::install_git("git://github.com/GabelHub/FAMoS.git", branch = "master")
```

## Features

### Adaptive methods of model selection

The FAMoS uses three different methods to find appropriate models to test:

- Forward search: Add a parameter to the currently best model
- Backward elimination: Remove a parameter from the currently best model
- Swap search: Replace a parameter of the currently best model by another. The parameters that can be swapped need to be specified by the user

The FAMoS keeps track of the methods used in the previous iterations and dynamically changes them according to the outcome of each iteration.

### Easy parallelisation

FAMoS makes use of the future-package which allows for easy parallelisation, meaning many different models can be tested simultaneously if the required computational resources are available. 

### Smart testing procedures

FAMoS keeps track of previously tested models and checks also that each model fulfills all user-specified restrictions, therefore testing only relevant models and saving computational resources.

## Example

As a simple example, we generate a simple data set generated by two parameters and apply the FAMoS on a global model consisting of five different parameters.

```{r example, echo=TRUE, eval = FALSE}
 library(FAMoS)
 #set the future plan (see future documentation for more information)
 future::plan(future::sequential)

 #create data with standard deviation of 1
 x.values <- 1:7
 y.values <-  9 * x.values^2 - exp(2 * x.values)
 sd.y.values <- rep(1,7)

 #define initial parameter values
 inits <- c(p1 = 3, p2 = 4, p3 = -2, p4 = 2, p5 = 0)

 #define cost function that returns the negative log-likelihood
 cost_function <- function(parms, x.vals, y.vals, sd.y){
  # restrict the search range to -5 to +5
  if(max(abs(parms)) > 5){
    return(NA)
  }
  with(as.list(c(parms)), {
    res <- p1*4 + p2*x.vals + p3^2*x.vals^2 + p4*sin(x.vals)  - exp(p5*x.vals)
    diff <- sum((res - y.vals)^2/sd.y)
  })
 }

 #perform model selection
 res <- famos(init.par = inits,
             fit.fn = cost_function,
             nr.of.data = length(y.values),
             init.model.type = c("p1", "p3"),
             optim.runs = 1,
             x.vals = x.values,
             y.vals = y.values,
             sd.y = sd.y.values)
```


The FAMoS returns a lot of verbose output, telling the user what's currently happening. In the beginning, the overall settings are defined and the corresponding directories are created (if they don't exist). 

```{r FAMoSstart, eval = FALSE}
#> Initializing...
#> Create FAMoS directory...
#> 
#> Algorithm run: 001
#> Information criterion for model comparison: AICc
#> Refitting disabled.
#> Starting algorithm with method 'forward'
```

In each iteration, the FAMoS identifies new models to be tested based on the current search method:

```{r iteration, eval = FALSE}
#> FAMoS iteration #2 - method: forward
#> Add p2
#> Add p4
#> Add p5
#> Time passed since start: 1.1 secs
```

Each model will be submitted and tested. Since FAMoS uses futures for evaluation, the search process can be easily parallelised by setting the corresponding future plan. Every model is subsequently evaluated by performing (multiple) optimisation routines based on _optim_.

```{r optim, eval = FALSE}
#> Job ID for model 01 - 11100:
#> 
#> Fitting run # 1
#>          p1          p2          p3 
#> -4.99979345  0.00000000  0.00570456 
#> 1471914833547.12 
#> 1471910549134.14 
#> 1471910349497.77 
#> 1471910340394.53 
#> 1471910340394.53 
#> No results file present. Current parameters saved.
#> 
#> Status file updated. Fitting done.
```

After all models have been evaluated, the algorithm reads in the results and checks, if a better model was found

```{r enditeration, eval = FALSE}
#> Evaluate results ...
#> Best AICc of this run is 14
#> Parameter p5 was added
#> Time passed since start: 1.92 secs
```

The cycle continues until no better model can be found based on the currently used methods. After halting, the results are returned

```{r bestresults, eval = F}
#> Best model found. Algorithm stopped.
#> FAMoS run 001
#> AICc of best model: 7
#> Best model (binary): 00101
#> Best model (vector):
#> p1 p2 p3 p4 p5 
#>  0  0  1  0  1 
#> Estimated parameter values:
#> p1 p2 p3 p4 p5 
#>  0  0 -3  0  2 
#> Time needed: 3.84 secs
```

## FAMoS options in detail

#### init.par

The vector _init.par_ is one of the three essential variables that need to be specified. It contains the names and initial values of _all_ parameters, that the FAMoS is supposed search through. In our example above, we specified this vector as

```{r init.par, eval = FALSE}
#define initial parameter values
 inits <- c(p1 = 3, p2 = 4, p3 = -2, p4 = 2, p5 = 0)
```

Depending on the starting model, FAMoS automatically extracts the corresponding values and uses them for its first iteration only. All following iterations inherit the best values found during fitting.

Additional specifications for the use of the inital parameter vector can be supplied by the options _do.not.fit_ and _default.val_. 

#### fit.fn
To allow independence of a certain mathematical model structure, the user can specify any cost function that is able to calculate the negative log-likelihood (-2LL) from a given parameter set. The necessity for returning the negative log-likelihood stems from the fact that model selection is performed on the AIC(c) or BIC. The basic structure of the cost function is therefore:

```{r fit.fn, eval = FALSE}
cost.function <- function(parms, ...){
  ... calculate the negative log-likelihood based on parms...
  return(negative log-likelihood)
}
```

Due to this structure, the FAMoS is able to tackle many different problems, e.g. modelling approaches based on ODEs or PDEs.

Note that the cost function needs to take the complete parameter set (as specified in _init.par_) as an input. The exact composition of the _parms_ vector is adaptively set by the FAMoS, which inherits the values to be fitted from the previous best fits and sets the non-fitted parameters to zero (or another value, if specified in the option _default.val_).

#### nr.of.data

To calculate the respective information criteria (AIC(c) or BIC, see _Burnham and Anderson 2002_ for details) the number of different data points needs to be supplied as well.

#### homedir

The FAMoS generates and saves many different files, in order to make results available over time as well as to simultaneously running FAMoS runs. _homedir_ specifies the folder, in which all results are going to be stored.

#### do.not.fit

In order to exclude some parameters from the fitting procedures, their names can be specified in the _do.not.fit_ option. This allows to test different model restrictions without needing to change either _init.par_ or _fit.fn_. For example, if we wanted to exlude the parameter _p4_ from our analysis, we would specify initially

```{r do.not.fit, eval = FALSE}
#define initial parameter values
 inits <- c(p1 = 3, p2 = 4, p3 = -2, p4 = 2, p5 = 0)
 no.fit <- c("p4")
```

and pass this option on to FAMoS. Note that excluded parameters are automatically removed from the initial model, if _init.model.type = "random"_ or _init.model.type = "global"_ is used. If the user-specfied initial model contains an exluded parameter, an error will be returned.

```{r donotfiterror, eval = FALSE}
  The specified initial model violates critical conditions or the do.not.fit specifications
```
#### method

FAMoS can use three different methods to search for different models to test: Forward search, backward elimination and swap search. As the algorithm dynamically changes these methods over the course of each iteration, the option _method_ only specifies the starting method. 

If the algorithm is able to find a better model, the current method will be used in the next iteration as well (except the swap method, which always uses a forward serch next - if it doesn't terminate in that step). If no better model is found, the algorithm will change the method according to the following scheme:

| current method  | previous method | next method |
| :-------------: |:-------------:| :-----:|
| forward      | backward | swap (or terminate)|
| forward      | forward or swap | backward |
| backward | backward |  forward |
| backward | forward |  swap (or terminate) |
| swap | forward or backward |  terminate |

In case the swap method is not used (due to unspecified critical or swap sets), the algorithm will terminate after a succession of an unsuccessful forward and backward search.

#### init.model.type

To verify if the FAMoS results are consistent, it is important to run the algorithm with different starting models. To set the initial model, the user can either use the built-in options _random_ (which generates a random model) or _global_ (which uses the complete model as a starting point). Alternatively, the user can specify a model by supplying a parameter vector containing the names of the initial model.

```{r init.model.type, eval = FALSE}
#Three options for the starting model
init.model1 <- "random" # generates a random starting model
init.model2 <- "global" # uses all available parameters
init.model3 <- c("p1", "p4") # a user-specified model
```

In case _random_ or _global_ are chosen, the FAMoS automatically applies critical conditions and removes excluded parameters (see options _critical.parameters_ and _do.not.fit_).

#### refit

Before testing a model, the FAMoS checks if this model has been tested before. In case _refit = FALSE_ (default) is specified, the model will not be tested again. If refitting is set to TRUE, FAMoS will try to optimise the model again. If the new run returns a better fit, the old results will be overwritten, otherwise the new run will be discarded.

Refitting makes sense if the model optimisation is dependent on the initial parameter combination (see also _optim.runs_). If a model is reencountered, it might well be that the new parameter set to be tested with is much more appropriate than the previous one, especially if this reencounter happens within the same FAMoS run.

#### optim.runs

Finding the best fit for each model is crucial to guarantee a correct model selection procedure. Fitting in FAMoS is based on the built-in function _optim_, which is repeatedly evaluated. Here, _optim.runs_ gives the number of fitting attempts with _different_ starting conditions. Here, the first fitting attempt takes the inherited parameter vectors from previous runs, while all following fitting attempts randomly samples parameter vectors to test.

As the default optimisation method is based on the Nelder-Mead approach, which - in my experience - tends to not give reliable results if only one optimisation is performed, the optimisation for each fitting attempt is wrapped into a while-loop, in which the fitting procedure is repeatedly halted and restarted (based on the options _control.optim_), until the absolute convergence tolerance in _con.tol_ is reached.

The skeleton of the underlying code looks like this:

```{r optim.runs, eval = FALSE}

for(i in 1:optim.runs){#number of fitting attempts specified by optim.runs
  start.parameters <- either the inherited or a randomly sampled set (for i > 2)
  
  while((optim.old - optim.new) < con.tol){
    ... run optim with start.parameters ...
    start.parameters <- new parameters estimated by optim
  }
}

```

#### information.criterion

The information criterion used for performing the model selection. Options include AICc, AIC or BIC (see Burnham and Anderson 2002 for more details).

#### default.val

Normally, FAMoS sets the parameters that are not fitted equal to zero. However, this might not be appropriate if, for example, a parameter describes an initial condition or a baseline turnover. Here,  _default.val_ allows to specify the value that a parameter assumes, if it is not fitted. _default.val_ needs to be given as a named list, which can either store numerical values or the name of the parameter from which the value should be inherited. For example

```{r default.val, eval = FALSE}
 #define initial parameter values
 inits <- c(p1 = 3, p2 = 4, p3 = -2, p4 = 2, p5 = 0)
 #set default values
 def.val <- list(p1 = 2, p2 = -5, p3 = "p1", p4 = 0, p5 = "p4")

```

Here, the values of _p1_, _p2_, and _p4_ are set to their respective values. However, _p3_ and _p5_ will inherit their values from _p1_ and _p4_, respectively. This feature is useful if two rates describe similar processes and one wants to test if the difference between them is significant enough to warrant the fitting of an additional parameter. Here's a short example

```{r defaultval2, eval = FALSE}
cost.function <- function(parms){
  x <- par1 + par2*x
  y <- par3 + par4*x
}

def.val <- list(p1 = 0, p2 = 0, p3 = "p1", p4 = "p2")
```

Note that the parameter inheritance cannot be chained, meaning that entries that point to another parameter need a numeric value to access

```{r defaultval3, eval = FALSE}
#INCORRECT use of default.val
def.val <- list(p1 = 1, p2 = "p1", p3 = "p2", p4 = "p3")
#CORRECT use of default.val
def.val <- list(p1 = 1, p2 = "p1", p3 = "p1", p4 = "p1")
```

#### swap.parameters

The swap search that FAMoS can perform relies on sets which specify parameters that can be swapped by one another. For example, if we wanted to allow parameters _p1_, _p2_ and _p3_, as well as _p4_ and _p5_ to be replaceable by each other, we would specify:

```{r swaps, eval = FALSE}

swap.set <- list(c("p1", "p2", "p3"), c("p4", "p5"))
```
#### critical.parameters

In some cases, it does not make sense to fit certain submodels of the global model to the data, as they might lack crucial parameters. FAMoS can incorporate these restrictions by the specification of critical parameter sets. For example, if at least one of the first three parameters need to be present in the model, and all models that don't feature _p4_ are not correct, we can specify:

```{r crits, eval = FALSE}
crit.set <- list(c("p1", "p2", "p3"), c("p4"))
```

All critical sets are also automatically used in the swap search.

#### random.borders

Since the parameters of all _optim.runs_ larger than one are sampled based on a random uniform distribution, it might be important to set the correct sampling intervals. By default, FAMoS samples parameters with a 100% deviation of the inherited parameter values (for example, if a model contains two parameters, and the currently best values are _p1 = 0.1_ and _p2 = -1000_, the sampled values will lie in the intervals [0,0.2] and [-2000,0], respectively). Alternatively, the user can specify relative or absolute sampling intervals. For relative intervals, a numeric value has to be given for each parameter denoting its relative deviation. For absolute sampling intervals, a matrix containing the lower and upper borders has to be specified. Here's an example:

```{r random.borders, eval = FALSE}
#relative sampling ranges
random.bord1 <- 0.3 # deviates all parameters by 30%
random.bord2 <- c(0.1, 0.5, 0.2) # deviates the parameters by 10%, 50% and 20%, respectively

#absolute sampling ranges
random.bord3 <- matrix(c(1,2), nrow = 1) #uses the interval [1,2] for all parameter samples
random.bord4 <- cbind(c(0,-10, 0.3), c(5, -9, 0.7)) #uses the intervals [0,5], [-10,-9] and [0.3, 0.7] to sample the respective parameters

```
#### control.optim

Specifies the control options used for _optim_ (see _optim.runs_ for more details).

#### con.tol

Specifies the absolute convergence tolerance used for _optim_ (see _optim.runs_ for more details).

#### save.performance

If true, a plot of the current FAMoS performance is stored in the folder "FAMoS-Results/Figures/", which will be updated during each iteration.
