---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# FAMoS

FAMoS provides an automated and unbiased model selection algorithm that aims at determining the most appropriate subset of model parameters to describe a specific data set. Due to its flexibility with respect to the cost/optimisation function, FAMoS can handle many different mathematical structures, including for example regression models and ODEs.

## Installation

You can install FAMoS from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("GabelHub/FAMoS")
# alternative installation command
devtools::install_git("git://github.com/GabelHub/FAMoS.git", branch = "master")
```

## Features

### Adaptive methods of model selection

FAMoS uses three different methods to find appropriate models to test:

- Forward search: Add a parameter to the currently best model
- Backward elimination: Remove a parameter from the currently best model
- Swap search: Replace a parameter of the currently best model by another. The parameters that can be swapped need to be specified by the user

FAMoS keeps track of the methods used in the previous iterations and dynamically changes them according to the outcome of each iteration.

### Flexibility

FAMoS is designed to allow for a maximum of flexibility regarding the fitting procedures and model types in R. While it comes with the default option to fit the cost function via _optim_, it also allows the users to specify their own optimisation routines, hence making it possible to perform model selection based on various other R packages.

### Easy parallelisation

FAMoS makes use of the future-package which allows for easy parallelisation, meaning many different models can be tested simultaneously if the required computational resources are available. 

### Smart testing procedures

FAMoS keeps track of previously tested models and checks also that each model fulfills all user-specified restrictions, therefore testing only relevant models and saving computational resources.

## Example

As a simple example, we generate a simple data set generated by two parameters and apply FAMoS on a global model consisting of five different parameters.

```{r example, echo=TRUE, eval = FALSE}
library(FAMoS)

#setting data
true.p2 <- 3
true.p5 <- 2
sim.data <- cbind.data.frame(range = 1:10, 
                             y = true.p2^2 * (1:10)^2 - exp(true.p5 * (1:10)))

#define initial parameter values and corresponding test function
inits <- c(p1 = 3, p2 = 4, p3 = -2, p4 = 2, p5 = 0)

cost_function <- function(parms, binary, data){
  if(max(abs(parms)) > 5){
    return(NA)
  }
  with(as.list(c(parms)), {
    res <- p1*4 + p2^2*data$range^2 + p3*sin(data$range) + p4*data$range - exp(p5*data$range)
    diff <- sum((res - data$y)^2)
    
    #calculate AICC
    nr.par <- length(which(binary == 1))
    nr.data <- nrow(data)
    AICC <- diff + 2*nr.par + 2*nr.par*(nr.par + 1)/(nr.data - nr.par -1)
    
    return(AICC)
  })
}


#set swap set
swaps <- list(c("p1", "p5"))

#perform model selection
res <- famos(init.par = inits,
             fit.fn = cost_function,
             homedir = tempdir(),
             method = "swap",
             swap.parameters = swaps,
             init.model.type = c("p1", "p3"),
             optim.runs = 1,
             data = sim.data) 
```


FAMoS returns a lot of verbose output, telling the user what's currently happening (Note: The output can be turned on and off by using the option _verbose_). In the beginning, the overall settings are defined and the corresponding directories are created (if they don't exist). 

```{r FAMoSstart, eval = FALSE}
#> Initializing...
#> Create FAMoS directory...
#> 
#> Algorithm run: 001
#> Refitting disabled.
#> Starting algorithm with method 'swap'
```

In each iteration, FAMoS identifies new models to be tested based on the current search method:

```{r iteration, eval = FALSE}
#> FAMoS iteration #3 - method: forward
#> Add parameter p1
#> Add parameter p2
#> Add parameter p4
#> Time passed since start: 00:00:00
```

Each model will be submitted and tested. Since FAMoS uses futures for evaluation, the search process can be easily parallelised by setting the corresponding future plan. Every model is subsequently evaluated by performing (multiple) optimisation routines based either on the default fitting routine _optim_ or a user-specified fitting routine (see the vignettes for examples).

After all models have been evaluated, the algorithm reads in the results and checks, if a better model was found

```{r enditeration, eval = FALSE}
#> Evaluate results ...
#> Best selection criterion value of this run is 10
#> Parameter p2 was added
#> Time passed since start: 1.92 secs
```

The cycle continues until no better model is found based on the currently used methods. After halting, the results are returned

```{r bestresults, eval = F}
#> Best model found. Algorithm stopped.
#> FAMoS run 001
#> Selection criterion value of best model: 7
#> Best model (binary): 01001
#> Best model (vector):
#> p1 p2 p3 p4 p5 
#>  0  1  0  0  1 
#> Estimated parameter values:
#> p1 p2 p3 p4 p5 
#>  0  -3 0  0  2 
#> Time needed: 3.84 secs
```

## FAMoS options in detail

#### init.par

The vector _init.par_ is one of two mandatory variables that need to be specified. It contains the names and initial values of _all_ model parameters, that FAMoS is supposed to analyse. In our example above, we specified this vector as

```{r init.par, eval = FALSE}
#define initial parameter values
inits <- c(p1 = 3, p2 = 4, p3 = -2, p4 = 2, p5 = 0)
```

Depending on the starting model, FAMoS automatically extracts the corresponding values and uses them for its first iteration only. All following iterations inherit the best values from previous fits.

Additional specifications for the use of the inital parameter vector can be supplied by the options _do.not.fit_ and _default.val_. 

#### fit.fn
To allow independence of specific mathematical model structures, the user can specify any cost or optimisation function. If a cost function is used, it has to take the complete parameter vector as an input (names _parms_) and has to return a selection criterion value. If use.optim = TRUE, the cost function needs to return a single numeric value, which corresponds to the selection criterion value. However, if use.optim = FALSE, the cost function needs to return a list containing in its first entry the selection criterion value and in its second entry the named vector of the fitted parameter values (non-fitted parameters are internally assessed).

Additionally, the cost and optimisation functions can also use the optional input _binary_, which contains the binary information of the current model, i.e. the information which parameters are currently considered to be fitted. This is useful to extract the to-be-fitted parameters, if a custom optimisation functions is used

Due to this flexible structure, FAMoS is able to tackle many different problems, e.g. modelling approaches like linear regression, ODEs or PDEs.

#### homedir

FAMoS generates and saves many different files, in order to make results available over time as well as to simultaneously running FAMoS runs. _homedir_ specifies the folder, in which all results are going to be stored. The default is set to the current working directory.

#### do.not.fit

In order to exclude some parameters from the fitting procedures, their names can be specified in the _do.not.fit_ option. This allows to test different model restrictions without needing to change either _init.par_ or _fit.fn_. For example, if we wanted to exclude the parameter _p4_ from our analysis, we would specify initially

```{r do.not.fit, eval = FALSE}
#define initial parameter values
inits <- c(p1 = 3, p2 = 4, p3 = -2, p4 = 2, p5 = 0)
no.fit <- c("p4")
```

and pass this option on to FAMoS. Note that excluded parameters are automatically removed from the initial model, if _init.model.type = "random"_, _init.model.type = "global"_ or  _init.model.type = "most.distant"_ is used. If the user-specified initial model contains an excluded parameter, an error will be returned.

```{r donotfiterror, eval = FALSE}
The specified initial model violates critical conditions or the do.not.fit specifications
```
#### method

FAMoS can use three different methods to search for different models to test: Forward search, backward elimination and swap search. As the algorithm dynamically changes these methods over the course of each iteration, the option _method_ only specifies the starting method. 

If the algorithm is able to find a better model, the current method will be used in the next iteration as well (except the swap method, which always uses a forward search next - if it doesn't terminate in that step). If no better model is found, the algorithm will change the method according to the following scheme:

| current method  | previous method | next method |
| :-------------: |:-------------:| :-----:|
| forward      | backward | swap (or terminate)|
| forward      | forward or swap | backward |
| backward | backward |  forward |
| backward | forward |  swap (or terminate) |
| swap | forward or backward |  terminate |

In case the swap method is not used (due to unspecified critical or swap sets), the algorithm will terminate after a succession of an unsuccessful forward and backward search.

#### init.model.type

To verify if FAMoS results are consistent, it is important to run the algorithm with different starting models. To set the initial model, the user can either use the built-in options _random_ (which generates a random model), _global_ (which uses the complete model as a starting point) or _most.distant_ (uses the model most dissimilar to all previously tested models). Alternatively, the user can specify a model by supplying a parameter vector containing the names of the initial model.

```{r init.model.type, eval = FALSE}
#Three options for the starting model
init.model1 <- "random" # generates a random starting model
init.model2 <- "global" # uses all available parameters
init.model3 <- "most.distant" # uses the most dissimilar model
init.model4 <- c("p1", "p4") # a user-specified model
```

In case _random_, _global_ or _most.distant_ are chosen, FAMoS automatically applies critical conditions and removes excluded parameters (see options _critical.parameters_ and _do.not.fit_).

#### refit

Before testing a model, FAMoS checks if this model has been tested before. In case _refit = FALSE_ (default) is specified, the model will not be tested again. If refitting is set to TRUE, FAMoS will try to optimise the model again. If the new run returns a better fit, the old results will be overwritten, otherwise the new run will be discarded.

Refitting makes sense if the model optimisation is dependent on the initial parameter combination (see also _optim.runs_). If a model is reencountered, it might well be that the new parameter set to be tested with is much more appropriate than the previous one, especially if this reencounter happens within the same FAMoS run.

#### use.optim

The default fitting routine that FAMoS relies on is the built-in function _optim_. However, by setting _use.optim = FALSE_, the user can use any other optimisation routine suitable. The optimisation routine then has to be included in the cost function _fit.fn_ which needs to return a list containing the current selection value criterion as well as the parameter values used. See the vignettes for an example.

#### optim.runs

Finding the best fit for each model is crucial to guarantee a correct model selection procedure. Often, fitting a model once is enough and repeating the fitting procedure with different initial conditions does not lead to new results. Sometimes, however, one wants to run multiple fits for each model, e.g. if the parameter space is very large. To do so, the user can specify _optim.runs_, which gives the number of fitting attempts. For each optimisation run a _different_ starting condition is used. The first fitting attempt takes the inherited parameter vectors from previous runs, while all following fitting attempts randomly samples parameter vectors to test (see also _random.borders_).

If multiple optimisation runs are performed, FAMoS will return the best of these runs.

In each optimisation run fitting in FAMoS is either performed with the built-in function _optim_, which is repeatedly evaluated until convergence, or a custom optimisation routine, which is evaluated only once. 
As the default optimisation method is based on the Nelder-Mead approach, which often tends to not give reliable results if only one optimisation is performed, the optimisation for each fitting attempt is wrapped into a while-loop, in which the fitting procedure is repeatedly halted and restarted (based on the options _control.optim_), until the relative convergence tolerance in _con.tol_ is reached.

The skeleton of the underlying code looks like this:

```{r optim.runs, eval = FALSE}

for(i in 1:optim.runs){#number of fitting attempts specified by optim.runs
  start.parameters <- either the inherited or a randomly sampled set (for i > 2)
  
  if(use.optim == TRUE){
    #If use.optim = TRUE, the fitting routine is evaluated in a while loop
    while(abs((old.optim.value - new.optim.value)/old.optim.value) < con.tol){
      ... run optim with start.parameters ...
      start.parameters <- new parameters estimated by optim
    }
  }else{
    #If use.optim = FALSE, the custom optimisation routine is evaluated 
    #only once in each optimisation run
    ... run custom optimisation with start parameters ...
  }
  
}

```

#### default.val

Normally, FAMoS sets the parameters that are not fitted equal to zero. However, this might not be appropriate if, for example, a parameter describes an initial condition or a baseline turnover. Here,  _default.val_ allows to specify the value that a parameter assumes, if it is not fitted. _default.val_ needs to be given as a named list, which can either store numerical values or the name of the parameter from which the value should be inherited. For example

```{r default.val, eval = FALSE}
#define initial parameter values
inits <- c(p1 = 3, p2 = 4, p3 = -2, p4 = 2, p5 = 0)
#set default values
def.val <- list(p1 = 2, p2 = -5, p3 = "p1", p4 = 0, p5 = "p4")

```

Here, the values of _p1_, _p2_, and _p4_ are set to their respective values. However, _p3_ and _p5_ will inherit their values from _p1_ and _p4_, respectively. This feature is useful if two rates describe similar processes and one wants to test if the difference between them is significant enough to warrant the fitting of an additional parameter. Here's a short example

```{r defaultval2, eval = FALSE}
cost.function <- function(parms){
  x <- par1 + par2*x
  y <- par3 + par4*x
}

def.val <- list(p1 = 0, p2 = 0, p3 = "p1", p4 = "p2")
```

Note that the parameter inheritance cannot be chained, meaning that entries that point to another parameter need a numeric value to access

```{r defaultval3, eval = FALSE}
#INCORRECT use of default.val
def.val <- list(p1 = 1, p2 = "p1", p3 = "p2", p4 = "p3")
#CORRECT use of default.val
def.val <- list(p1 = 1, p2 = "p1", p3 = "p1", p4 = "p1")
```

#### swap.parameters

The swap search that FAMoS can perform relies on sets which specify parameters that can be swapped by one another. For example, if we wanted to allow parameters _p1_, _p2_ and _p3_, as well as _p4_ and _p5_ to be replaceable by each other, we would specify:

```{r swaps, eval = FALSE}

swap.set <- list(c("p1", "p2", "p3"), c("p4", "p5"))
```
#### critical.parameters

In some cases, it does not make sense to fit certain submodels of the global model to the data, as they might lack crucial parameters. FAMoS can incorporate these restrictions by the specification of critical parameter sets. For example, if at least one of the first three parameters need to be present in the model, and all models that don't feature _p4_ are not correct, we can specify:

```{r crits, eval = FALSE}
crit.set <- list(c("p1", "p2", "p3"), c("p4"))
```

All critical sets are also automatically used in the swap search.

#### random.borders

Since the parameters of all _optim.runs_ larger than one are sampled based on a random uniform distribution, it might be important to set the correct sampling intervals. By default, FAMoS samples parameters with a 100% deviation of the inherited parameter values (for example, if a model contains two parameters, and the currently best values are _p1 = 0.1_ and _p2 = -1000_, the sampled values will lie in the intervals [0,0.2] and [-2000,0], respectively). Alternatively, the user can specify relative or absolute sampling intervals. For relative intervals, a numeric value has to be given for each parameter denoting its relative deviation. For absolute sampling intervals, a matrix containing the lower and upper borders has to be specified. Here's an example:

```{r random.borders, eval = FALSE}
#relative sampling ranges
random.bord1 <- 0.3 # deviates all parameters by 30%
random.bord2 <- c(0.1, 0.5, 0.2) # deviates the parameters by 10%, 50% and 20%, respectively

#absolute sampling ranges
random.bord3 <- matrix(c(1,2), nrow = 1) #uses the interval [1,2] for all parameter samples
random.bord4 <- cbind(c(0,-10, 0.3), c(5, -9, 0.7)) #uses the intervals [0,5], [-10,-9] and [0.3, 0.7] to sample the respective parameters

#use a function to sample the results
random.bord5 <- rnorm #note that in this case, 'mean' and 'sd' need to passed to famos as well, if other values than the default settings should be used

```
#### control.optim

Specifies the control options used for _optim_ (see _optim.runs_ for more details).

#### parscale.pars
If parameters values span over several orders of magnitudes, using the built-in option _parscale_ in _optim_ can reduce the numbers of evaluations needed. Setting _parscale.pars = TRUE_ automatically adjusts the scaling procedure in _optim_. In our experience, using _parscale.pars = TRUE_ is usually beneficial if a large number of parameters with different orders of magnitude need to be fitted. However, the actual performance is very problem-specific and therefore we would recommend initially testing both approaches to see which one performs better for the problem at hand. Also, one needs to make sure that the other options given in _control.optim_ and _con.tol_ are specified appropriately.

#### con.tol

Specifies the relative convergence tolerance and determines when the repeated use _optim_ fits will be terminated (see _optim.runs_ for more details).  

#### save.performance

If true, a plot of the current FAMoS performance is stored in the folder "FAMoS-Results/Figures/", which will be updated during each iteration.

#### use.futures

To allow for parallelisation, FAMoS uses the _future_ package by Henrik Bengtsson (see https://github.com/HenrikBengtsson/future). To use futures, the option needs to be set to TRUE and a future plan needs to be specified.

#### reattempt

By default, FAMoS terminates once all search methods are exhausted. However, if _reattempts_ is set to true, FAMos will instead jump to a distant model and continues to search the model space from there. The algorithm is then terminated if the best model is re-encountered (or if no other models are available to test).

#### log.interval

If futures are used during a FAMoS run, there will be a message printed every _X_ seconds, informing the user which models fits are still running. _log.interval_ allows to specify the interval of _X_. Default to 10 minutes (600 seconds).

#### interactive.session

As FAMos allows to use previously generated results in later runs, it performs a consistency check to see if the previous results were generated by the same cost function. If this is not the case, FAMoS requires user input to decide what to do next. However, if run non-locally, supplying user input might not b possible. Therefore, *interactive.session* can be set to FALSE. This will result in FAMoS issuing a warning instead of an interaction prompt.

#### verbose

The verbose output of FAMoS can be turned on and off. If _verbose = FALSE_, only a minimum of information is shown.
